Partial rank data with the \pkg{hyper2} package: likelihood functions for generalized Bradley-Terry models

### 2019-05-06 inst/collusion_chess_draws.Rnw
work

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{booktabs}

@@

We see that $\widehat{p_D}\simeq 0.53$, compare the naive estimate of
$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
matches.  Secondly, we may make inferences about $p_W$.  It is
conceivable that $p_W=0$ as this term does not appear on its own in
the numerator of any term.  We may test $H_0\colon p_W=0$ using the
method of support.

@@

that is, well over 7 units of support, exceeding Edwards's
two-units-of-support per degree of freedom and we can reject $p_W=0$.

One very natural hypothesis would be $H_{A=K=K}\colon
p_\mathrm{Karpov}=p_\mathrm{Kasparov}=p_\mathrm{Anand}$.  This too is
readily testable:

@@

>>>>

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{booktabs}
\usepackage{units}

@@

We see that $\widehat{p_D}\simeq 0.53$, compare the na\"ive estimate
of $\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
matches.  Secondly, we may make inferences about $p_W$.  It is
conceivable that $p_W=0$ as this term does not appear on its own in
the numerator of any term.  We may test $H_0\colon p_W=0$ using the
method of support.

@@

that is, well over 7 units of support, exceeding Edwards's
two-units-of-support per degree of freedom and we can reject $p_W=0$.
One very natural hypothesis would be $H_{A=Kr=Ks}\colon
p_\mathrm{Karpov}=p_\mathrm{Kasparov}=p_\mathrm{Anand}$.  This too is
readily testable:

<<>>=
objective <- function(V){
  x <- V[1]  # strength of all three players
  w <- V[2]  # white player advantage
  D <- 1-3*x-w  # Draw monster
  return(  
      -loglik(karpov_kasparov_anand,c(x,x,x,w))  # negative as we minimize
  )
}

jj <-
constrOptim(
    theta=c(0.1,0.4),
    f=objective,
    grad=NULL,
    ui = rbind(diag(2),-diag(2),c(-3,-1)),
    ci = c(0,0,-1,-1,-1)
)
jj

(unconstrained <- maxp(karpov_kasparov_anand,give=TRUE)$value)
(constrained <- jj$value)

(extra_support <-   unconstrained + constrained)  # negative used

pchisq(2*extra_support,df=2,lower.tail=FALSE)
@

highly significant.  We can justifiably assert that the players do
indeed have different strengths, even if we account for white
advantage and the proclivity to draw.

\subsection{Each player has a draw monster}

<<>>=
objective <- function(V){
  karpov <- V[1]
  kasparov <- V[2]
  anand <- V[3]
  white <- V[4]

  draw <- (1-karpov-kasparov-anand-white)/3

    return(  
        -loglik(kka_3draws,
                c(karpov,kasparov,anand,white,draw,draw)
                )
        )
}

jj <-
constrOptim(
    theta=c(0.1,0.1,0.1,0.1),
    f=objective,
    grad=NULL,
    ui <- rbind(
        diag(nrow=4),
        c(-1,-1,-1,-1)
        ),
    ci = c(
        rep(0,4),
        -1
        )
)
jj

(unconstrained <- maxp(kka_3draws,give=TRUE)$value)
(constrained <- jj$value)

(extra_support <-   unconstrained + constrained)

pchisq(2*extra_support,df=2,lower.tail=FALSE)

@

Significant according to Edwards but not Fisher!

\section{Chess collusion}

@@

\newcommand{\half}{$\nicefrac{1}{2}$}

\begin{table}
\centering
\begin{tabular}{l|ccccc}
& \rotatebox{-90}{Fisher} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian} 
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip} 
\\ \hline
  Fisher       & - & \half&\half& 1&\half\\
Geller &\half&-&\half&\half&\half\\
Petrosian &\half&\half&-&\half&\half\\
Korchnoi &0&\half&\half&-&1\\
Filip &\half&\half&\half&0\\
\end{tabular}
\end{table}
